---
title: 'Review for Exam 1'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    incremental: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, echo = F, results = 'hide', message=FALSE}
library(tidyverse)
library(broom)
library(palmerpenguins)
penguins = penguins
```

### Review Problem 1a

```{r problem-1a, echo = F}
summary(lm(body_mass_g ~ species, data = penguins))
```

- Write the corresponding equation
- Interpret all coefficients as well as interpret any NHST procedures
- Interpret the overall model. Is it a good model? Why/why not?
- What is the mean body mass for each penguin type? 

---
### Review Problem 1b

```{r problem-1b, echo = F}
summary(aov(body_mass_g ~ species, data = penguins))
```

What are the similarities and differences between this output and the output on the previous page?

---

### Review Problem 2a

```{r problem-2a, echo = F}
summary(lm(body_mass_g ~ bill_length_mm, data = penguins))
```

- Write the corresponding equation
- Interpret all coefficients as well as interpret any NHST procedures
- Interpret the overall model. Is it a good model? Why/why not?

---

### Review Problem 2b

```{r problem-2b, echo = F}
summary(lm(body_mass_g ~ bill_length_mm + flipper_length_mm, data = penguins))
```

- Interpret all coefficients
- What are the similarities and differences between the previous model (2a) and the current model (2b)? Which is better?

---

### Review Problem 2c

```{r problem-2c, echo = F}
zPenguins = as.data.frame(scale(penguins[,c(3,5,6)]))
summary(lm(body_mass_g ~ bill_length_mm + flipper_length_mm, data = zPenguins))
```

- What are the similarities and differences between the previous model (2a) and the current model (2b)? Which is better?
- Interpret all coefficients

---

### Review Problem 3

.pull-left[
```{r problem-3, warning = FALSE, message = FALSE, echo=FALSE}
model = lm(body_mass_g ~ bill_length_mm + flipper_length_mm, data = penguins)
model = augment(model)

ggplot(data = model, aes(x = .fitted, y = .resid)) +
  geom_point(color = "cornflowerblue") +
  geom_smooth(color = "black", method = "lm") +
  theme_classic(base_size = 20)

```
]

.pull-right[
- This plot corresponds to the model in problem 2b
- Based on this, what assumptions can we check?
- Did this model satisfy the assumptions?
]

---

### Review Problem 4

- What is multiple R correlation?

- What is $R^2$?

- What is Adjusted $R^2$?

- What is tolerance?

- What does a semi-partial correlation tell you? What about a partial correlation?
Why do we care about either of these?



