---
title: 'Review for Exam 2'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    incremental: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, echo = F, results = 'hide', message=FALSE}
library(tidyverse)
library(broom)
library(palmerpenguins)
penguins = penguins

penguins$group = rep(c("group 1", "group 2", "group 3"), length.out = nrow(penguins))
penguins$group = factor(penguins$group)
```

### Review Question 1

```{r}
q1Mod = lm(body_mass_g ~ group * island, data = penguins)
summary(q1Mod)
```

---
### Review Question 1

```{r}
anova(q1Mod)
```
---
### Review Question 1

- Write the regression equation
- Interpret the lower order terms
- Fill in the following table of cell means
- If I blocked out pieces of information from the ANOVA table, could you fill them in? How do you interpret the ANOVA table findings?


|             | Biscoe | Dream | Torgersen |
|-------------|--------|-------|-----------|
| **Group 1** |        |       |           |
| **Group 2** |        |       |           |
| **Group 3** |        |       |           |

---

### Review Question 2

```{r}
summary(lm(body_mass_g ~ bill_length_mm * flipper_length_mm, data = penguins))
```

---

### Review Question 2

- Write the regression equation
- Interpret all terms
- Interpret omnibus NHST and omnibus effect size 
- What changes if you center your variables? What if you $z$-score them? Would you be able to recognize these transformations were done? 

---

### Other things to keep in mind

- DAGs are on the exam! Can you draw one? Can you recognize the 3 different ways we transmit information? How does "controlling for" variables matter in the context of DAGs? 
- Do you know when to use a polynomial regression? How would you know if your current regression should actually include a polynomial term?
- Can you draw your interactions? What types of things should be included? 

---
class: inverse, center, middle

# DO NOT LOOK UNTIL AFTER YOU HAVE TRIED THE REVIEW QUESTIONS!