---
title: 'Univariate regression'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "my-theme.css"]
    incremental: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 4em 1em 4em;
}

.small .remark-code { 
  font-size: 80% !important;
}
.tiny .remark-code {
  font-size: 50% !important;
}

</style>

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#17139C",link_color = "#DD3E3E")

```

```{r, echo = F, warning = F, message = F, results = 'hide'}
library(psych)
library(tidyverse)
library(broom)
```


## Last time

- Correlation as inferential test

- Fisher's r to z transformation

- Correlation matrices

- Interpreting effect size

---

## Today

**Regression**

- What is it? Why is it useful

- Nuts and bolts

  - Equation

  - Ordinary least squares

  - Interpretation
  
---

## GLM/Regression

GLM is a general data analytic system, meaning lots of things fall under the umbrella of regression.

* *General* - broad set of similar models; can be applied to almost any context
  * IVs can be continuous, categorical, nominal, ordinal....
* *Linear* - We try to understand our dependent variable (DV) via a linear combination predictor variables (add & multiply)

---
## GLM/Regression
#### What do we get?
- effect sizes
- statistical significance
- incorporate multiple IVs
- account for intercorrelations

---

### Regression

- **Scientific** use: explaining the influence of one or more variables on some outcome. 

  
- **Prediction** use: We can develop models based on what's happened in the past to predict what will happen in the figure.

- **Adjustment**: Statistically control for known effects

  + If everyone had the same level of SES, would abuse still be associated with criminal behavior?

---

## Regression equation

What is a regression equation?

- Functional relationship

  - Ideally like a physical law $(E = MC^2)$
  
  - In practice, it's never as robust as that
  
How do we uncover the relationship?

---

### How does Y vary with X?


- $\large E(Y|X)$

- "Our best guess" regardless of whether our model includes categories or continuous predictor variables

- We will evaluate our guesses based on how far away we are from the mean. But how do we come up with those guesses in the first place?

---

## Regression Equation

$$\Large Y_i = b_{0} + b_{1}X_i +e_i$$

$$\Large \hat{Y_i} = b_{0} + b_{1}X_i$$

$\hat{Y}$ signifies the predicted score -- no error

The difference between the predicted and observed score is the residual ( $e_i$ )

There is a different $e$ value for each observation in the dataset

---

## OLS
- How do we find the regression estimates? 
- Ordinary Least Squares (OLS) estimation
- Minimizes deviations 

$$ min\sum(Y_{i}-\hat{Y})^{2} $$ 

- Other estimation procedures possible (and necessary in some cases)

---

 
```{r,echo=FALSE, message=FALSE, cache=TRUE, warning = F}
set.seed(123)
x.1 <- rnorm(10, 0, 1)
e.1 <- rnorm(10, 0, 2)
y.1 <- .5 + .55 * x.1 + e.1
d.1 <- data.frame(x.1,y.1)
m.1 <- lm(y.1 ~ x.1, data = d.1)
d1.f<- augment(m.1)
```

```{r plot1, echo=FALSE, cache=TRUE, eval = T, warning=FALSE, message=FALSE}
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) +
  theme_bw(base_size = 20)
```

---

```{r, echo=FALSE}
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  theme_bw(base_size = 20)
```


---

```{r, echo=FALSE}
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = .fitted))+
  theme_bw(base_size = 20)
```



---

```{r, echo=FALSE, wanring = FALSE, message = FALSE}
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = .fitted))+
  theme_bw(base_size = 20)
```


---

```{r, echo = F}
new.i = 1.1
new.slope = -0.7
d1.f$new.fitted = 1.1 -0.7*d1.f$x.1

ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_abline(intercept = new.i, slope = new.slope, color = "blue", linewidth = 1) +
  geom_point(aes(y = new.fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = new.fitted))+
  theme_bw(base_size = 20)
```

## compare to bad fit
---

## What is error?

$$\Large Y_i = b_{0} + b_{1}X_i +e_i$$

$$\Large \hat{Y_i} = b_{0} + b_{1}X_i$$

$$\Large Y_i = \hat{Y_i} + e_i$$

$$\Large e_i = Y_i - \hat{Y_i}$$

---

## OLS

The line that yields the smallest sum of squared deviations

$$\Large \Sigma(Y_i - \hat{Y_i})^2$$
$$\Large = \Sigma(Y_i - (b_0+b_{1}X_i))^2$$
$$\Large = \Sigma(e_i)^2$$

--

In order to find the OLS solution, you could try many different coefficients $(b_0 \text{ and } b_{1})$... or not

---
## Regression coefficient, $b_{1}$

$$\large b_{1} = \frac{cov_{XY}}{s_{x}^{2}} = r_{xy} \frac{s_{y}}{s_{x}}$$

$$\large r_{xy} = \frac{s_{xy}}{s_xs_y}$$


*Suggested Practice: Go in `R` and see if you can prove to yourself that these equations are the same!*

---
## Standardized regression

- Regression using z-scores for Y and X
- Correlation equals standardized regression coefficient

$$\large b_{1} = r_{xy} \frac{s_{y}}{s_{x}}$$

$$ \large r_{xy} = b_1\frac{s_x}{s_y} $$
---
## Standardized regression

If the variance of both X and Y is equal to 1 (as in z-scores): 

$$\large \beta_{1} = b_{1}^* = r_{xy}$$

---

## Standardized regression equation

$$\large Y = b_{1}^*X+e$$

$$\large b_{1}^* = b_{1}\frac{s_x}{s_y}$$
--

According to this regression equation, when $X = 0, Y = 0$. Our interpretation of the coefficient is that a one-standard deviation increase in X is associated with a $b_{1}^*$ standard deviation increase in Y. Our regression coefficient is equivalent to the correlation coefficient *when we have only one predictor in our model.*

---

## Estimating the intercept raw $b_0$

- Re-write equation to include $\bar{X}$ & $\bar{Y}$
- Intercept serves to adjust for differences in means between X and Y

$$\hat{Y} = \bar{Y} + r_{xy} \frac{s_{y}}{s_{x}}(X-\bar{X})$$
- If standardized, intercept drops out
- Otherwise, intercept is where regression line crosses the y-axis at X = 0  
- When  $X = \bar{X}$ the regression line goes through  $\bar{Y}$. This is true for all regressions -- line must pass through $\bar{X}$  and  $\bar{Y}$

---



```{r}
galton.data <- psychTools::galton
head(galton.data)
describe(galton.data, fast = T)
cor(galton.data)
```

---

If we regress child height (Y) onto parents' (X):

```{r}
r = cor(galton.data)[2,1]
m_parent = mean(galton.data$parent)
m_child = mean(galton.data$child)
s_parent = sd(galton.data$parent)
s_child = sd(galton.data$child)

(b1 = r*(s_child/s_parent))
(b0 = m_child - b1*m_parent)
```


How will this change if we regress parent height onto child height?

---

```{r}
(b1 = r*(s_child/s_parent))
(b0 = m_child - b1*m_parent)
```


```{r}
(b1 = r*(s_parent/s_child))
(b0 = m_parent - b1*m_child)
```

---
## In `R`

```{r}
fit.1 <- lm(child ~ parent, data = galton.data)
summary(fit.1)
```

---
## Reversed

```{r}
summary(lm(parent ~ child, data = galton.data))
```

???

**Things to discuss**

- Coefficient estimates
- Statistical tests (covered in more detail soon)

---


```{r, echo=FALSE, cache=TRUE, warning = FALSE, message = FALSE}
ggplot(galton.data, aes(x=parent, y=child)) +
    geom_point() +    
    geom_smooth(method=lm,   # Add linear regression line
                se=FALSE) +
  theme_bw(base_size = 20)
```

---

### Data, predicted, and residuals

```{r}
library(broom)
model_info = augment(fit.1)
head(model_info)
```

```{r}
describe(model_info)
```

???

Point out the average of the residuals is 0, just like average deviation from the mean is 0. 

---

## Residuals

- Dispersion of residuals can be thought of as what is left over in Y that is *not* explained by our model. As residuals get smaller on average, so will the SD of the residuals.

- Sigma ( $\sigma$ ) is the SD of residuals. It can be thought of as how much left over in Y that we cannot explain by our model.

---

```{r, warning=FALSE, message=FALSE}
model_info %>% ggplot(aes(x = parent, y = .fitted)) +
  geom_point() + geom_smooth(se = F, method = "lm") + ggtitle(expression(paste("X is related to ", hat(Y))))+
  scale_x_continuous("X") + scale_y_continuous(expression(hat(Y))) + theme_bw(base_size = 30)
```

---

```{r, warning=FALSE, message=FALSE}
model_info %>% ggplot(aes(x = parent, y = .resid)) +
  geom_point() + geom_smooth(se = F, method = "lm") + ggtitle("X is always unrelated to e")+
  scale_x_continuous("X") + scale_y_continuous("e") + theme_bw(base_size = 30)
```

---

```{r, warning=FALSE, message=FALSE}
model_info %>% ggplot(aes(x = child, y = .fitted)) +
  geom_point() + geom_smooth(se = F, method = "lm") + ggtitle(expression(paste("Y can be related to ", hat(Y))))+
  scale_x_continuous("Y") + scale_y_continuous(expression(hat(Y))) + theme_bw(base_size = 30)
```

---

```{r, warning=FALSE, message=FALSE}
model_info %>% ggplot(aes(x = child, y = .resid)) +
  geom_point() + geom_smooth(se = F, method = "lm") + ggtitle("Y is sometimes related to e")+
  scale_x_continuous("Y") + scale_y_continuous("e") + theme_bw(base_size = 25)
```

---

```{r, warning=FALSE, message=FALSE}
model_info %>% ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() + geom_smooth(se = F, method = "lm") + ggtitle(expression(paste(hat(Y), " is always unrelated to e")))+
  scale_y_continuous("e") + scale_x_continuous(expression(hat(Y))) + theme_bw(base_size = 30)
```

---

```{r, warning=FALSE, message=FALSE}
model_info %>% rename(y = child, x = parent) %>% select(x,y,.fitted,.resid) %>% gather("key", "value") %>%
  ggplot(aes(value, fill = key)) + geom_histogram(bins = 25) + guides(fill = F)+
  facet_wrap(~key, scales = "free") + theme_bw(base_size = 20)
```

---

## Residuals Summary

- Residuals are not correlated with $X$ and $\hat{Y}$ because those two are perfectly correlated with one another (that is, $r_{\text{fitted,x}} = 1$  )
- $X$ and $\hat{Y}$ represent the *same* information. We use our model ( $X$ ) to make a prediction ( $\hat{Y}$ ).
- No correlation between residuals with $X$ and $\hat{Y}$ because they are created by subtracting them out of $Y$. ( $\epsilon = Y - \hat{Y}$ )
- $\sigma$  (SD of residuals) can be thought of as how much left over in $Y$ after we take out all of the information our model provides. 

---

class: inverse

## Next time... 

Statistical inferences with regression