---
title: 'Review for Exam 2'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    incremental: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, echo = F, results = 'hide', message=FALSE}
library(tidyverse)
library(broom)
library(palmerpenguins)
penguins = penguins

penguins$group = rep(c("group 1", "group 2", "group 3"), length.out = nrow(penguins))
penguins$group = factor(penguins$group)
```

### Tips 
- Prepare as though this were a closed note test that lasts 80 minutes. Do not rely on open notes. 

- Make a cheat sheet (preferably hand written) to keep next to you the entire time.

- Consider printing the exam and doing it by hand. Allows you to implement classic test taking strategies like highlighting/underlining etc.

- Study with people. Study with different people! 

- Speak out loud. You are learning a language. You don't become fluent in French without speaking French...

---
### Review Question 1

```{r, echo=FALSE}
q1Mod = lm(body_mass_g ~ group * island, data = penguins)
```

```{r, echo=FALSE}
summary(lm(body_mass_g ~ group * island, data = penguins))
```

---
### Review Question 1

```{r}
anova(q1Mod)
```
---
### Review Question 1

- Write the regression equation
- Interpret the lower order terms
- Fill in the following table of cell means
- If I blocked out pieces of information from the ANOVA table, could you fill them in? How do you interpret the ANOVA table findings?


|             | Biscoe | Dream | Torgersen |
|-------------|--------|-------|-----------|
| **Group 1** |        |       |           |
| **Group 2** |        |       |           |
| **Group 3** |        |       |           |

---

### Review Question 2

```{r}
summary(lm(body_mass_g ~ bill_length_mm * flipper_length_mm, data = penguins))
```

---

### Review Question 2

- Write the regression equation
- Interpret all terms
- Interpret omnibus NHST and omnibus effect size 
- What changes if you center your variables? What if you $z$-score them? Would you be able to recognize these transformations were done? 

---

### Other things to keep in mind

- Josh's bayesian lecture is on the exam! I only expect you to understand at a very high level what taking a bayesian approach might mean.

- Do you know when to use a polynomial regression? How would you know if your current regression should actually include a polynomial term?

- Can you draw your interactions? What types of things should be included? 

---
class: inverse, center, middle

# DO NOT LOOK UNTIL AFTER YOU HAVE TRIED THE REVIEW QUESTIONS!

---

```{r, warning=FALSE, message=FALSE}
means = penguins %>% 
  group_by(group, island) %>% 
  summarize(mean(body_mass_g, na.rm = T))

means
```

---
class:center, middle
|             |  Biscoe  |  Dream   | Torgersen |
|-------------|:--------:|:--------:|:---------:|
| **Group 1** | 4661.818 | 3720.349 | 3776.562  |
| **Group 2** | 4699.545 | 3761.310 | 3732.353  |
| **Group 3** | 4784.211 | 3652.564 | 3619.444  |
